---
title: "AdditionalAnalysis"
author: "d-muenter"
output: html_document
---

Because there are 985 different types of events in the data, we only look at those events that make up the majority of all the observations.  
We can see that only 30 different types of events make up almost 99 % of all the observations.
```{r}
## Find number of different types of events
length(unique(data$EVTYPE))
## Turn event variable into a factor variable
data$EVTYPE <- as.factor(data$EVTYPE)
## Reorder that variable by the frequency of each factor level
data$EVTYPE <- reorder(data$EVTYPE, data$EVTYPE, length)
## Look at the 30 most frequent event types
tail(table(data$EVTYPE), n=40)
sum(tail(table(data$EVTYPE), n=40)) / length(data$EVTYPE)
## Create a new smaller dataset with only the 30 most frequent event types
smalldata <- data[data$EVTYPE %in% names(tail(table(data$EVTYPE), n=40)),]
```
To make sure that by taking only the 40 most frequent types of events, we don't miss any relevant events which only occured a few times but caused significant health or economic effects, we take a look at which single events caused the most fatalities, injuries, property damage and crop damage.
```{r}
## What were the most fatalities in any of the recorded events?
mostfatalities <- head(sort(data$FATALITIES, decreasing=T), n=10)
mostfatalities
## To which type of event do these events correspond?
fatalityevents <- data$EVTYPE[data$FATALITIES %in% mostfatalities]
fatalityevents
## Are these types in our 40 selected types of events?
fatalityevents %in% smalldata$EVTYPE
```



```{r}
library(ggplot2)
fatal_plot <- ggplot(healthdata, aes(EVTYPE, FATALITIES)) + 
    geom_col(aes(fill=EVTYPE)) + guides(fill=FALSE) + 
    theme(axis.text.x=element_text(angle=45, hjust=1)) + 
    labs(x="Event Type", y="Fatalities")
injur_plot <- ggplot(healthdata, aes(EVTYPE, INJURIES)) + 
    geom_col(aes(fill=EVTYPE)) + guides(fill=FALSE) + 
    theme(axis.text.x=element_text(angle=45, hjust=1)) + 
    labs(x="Event Type", "Injuries")
library("gridExtra")
grid.arrange(fatal_plot, injur_plot, nrow=2)


ggplot(healthdata) + geom_col(aes(EVTYPE, FATALITIES)) + geom_col(aes(EVTYPE, INJURIES)) + theme(axis.text.x=element_text(angle=45, hjust=1)) + scale_y_continuous(sec.axis=sec_axis(trans=~.*1))
```

